### 2017-04-05 doc/Thesis/data/abstract.tex
Thesis: Changes after OGAPS round 1 feedback

\indent Transient reactor simulations are a chronically formidable challenge due to the computational rigor of evaluating the neutron transport equation, as well as its coupling with other physical phenomena. The goal of this thesis research is to mitigate the computational expense of these simulations by investigating and developing the improved quasi-static method (IQS). IQS is a rigorous space/time multiscale approach whereby the neutron flux is represented by a time-dependent amplitude and a time-space-energy dependent shape.  The objective of the IQS factorization is to evaluate amplitude and shape on different time scales in order to reduce computational burden associated with solving the multi-dimensional flux equations, while maintaining solution accuracy. IQS factorization leads to a nonlinear system of equations that requires iteration of shape and amplitude. Furthermore, reactor simulations often require coupling to additional multiphysics components for temperature feedback. These additional physics may evolve on different time scales and the applicability of IQS in multiphysic simulations needs to be evaluated from this perspective as well.

The objectives of this research is to establish IQS performance with various iteration techniques, validate time step convergence of IQS, and apply IQS to multiphysics simulation. IQS iteration techniques involve fixed-point (Picard) iteration with various convergence criteria and Newton iteration, namely preconditioned Jacobian-free Newton Krylov (PJFNK) method. Nonlinear convergence of each of these techniques is investigated. Validation of IQS with analysis of time step convergence is vital for implementation of time adaptive methods and error prediction. The time derivative of the shape function is discretized through fourth order using implicit-Euler, Crank-Nicolson, backward difference formulae (BDF), and singly-diagonally-implicit Runge-Kutta (SDIRK) methods. IQS application to multiphysics simulations involves its implementation into the Rattlesnake/MOOSE framework. These simulations allow insight to the performance of IQS for full transient reactor simulations.

>>>>

\indent Transient reactor simulations are a chronically formidable challenge due to the computational rigor of evaluating the neutron transport equation, as well as its coupling with other physical phenomena. The goal of this thesis research is to mitigate the computational expense of these simulations by investigating and developing the improved quasi-static method (IQS). IQS is a rigorous space/time multiscale approach whereby the neutron flux is represented by a time-dependent amplitude and a time-space-energy dependent shape.  The objective of the IQS factorization is to evaluate amplitude and shape on different time scales in order to reduce computational burden associated with solving the multi-dimensional flux equations, while maintaining solution accuracy. IQS factorization leads to a nonlinear system of equations that requires iteration of shape and amplitude. Furthermore, reactor simulations often require coupling to additional multiphysics components for temperature feedback. These additional physics may evolve on different time scales, and the applicability of IQS in multiphysic simulations needs to be evaluated from this perspective as well.

The objectives of this research are to establish IQS performance with various iteration techniques, validate time step convergence of IQS, and apply IQS to multiphysics simulation. IQS iteration techniques involve fixed-point (Picard) iteration with various convergence criteria and Newton iteration, namely preconditioned Jacobian-free Newton Krylov (PJFNK) method. Nonlinear convergence of each of these techniques is investigated. Validation of IQS with analysis of time step convergence is vital for implementation of time adaptive methods and error prediction. The time derivative of the shape function is discretized through fourth order using implicit-Euler, Crank-Nicolson, backward difference formulae (BDF), and singly-diagonally-implicit Runge-Kutta (SDIRK) methods. IQS application to multiphysics simulations involves its implementation into the Rattlesnake/MOOSE framework. These simulations allow insight into the performance of IQS for full transient reactor simulations.

### 2017-04-05 doc/Thesis/data/theory.tex
Thesis: Changes after OGAPS round 1 feedback

\chapter{Improved Quasi-Static Method}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory}
\label{sect:theory}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

To derive the amplitude equation, the shape/precursors equations are weighted by a space-dependent function and integrated over the phase-space. The weight function is typically the adjoint flux $\phi^{*g}$ at the initial time, which can be proven to minimize truncation error \cite{duderstadt1976nuclear}. The final expressions are given below:
\begin{subequations}
\be
\frac{dp}{dt}=\left[\frac{\rho-\bar{\beta}}{\Lambda}\right]p+\sum_{i=1}^I\bar{\lambda}_i\xi_i
\label{eq:p}
\ee
\be
\frac{d\xi_i}{dt}=\frac{\bar{\beta}_i}{\Lambda}-\bar{\lambda}_i\xi_i \quad 1 \le i \le I 
\label{eq:c}
\ee
\end{subequations}
This equation is also known as the point reactor kinetics equation (PRKE); where the reactivity, effective delayed-neutron fraction, and delayed-neutron precursor decay constant are defined as follows:
\begin{subequations}
\be
\frac{\rho-\bar{\beta}}{\Lambda} = 
\frac{ \sum_{g=1}^G\left(\phi^{*g},\frac{\chi_p^g}{\keff}(1-\beta)\sum_{g'=1}^G \nu^{g'} \Sigma_f^{g'}\varphi^{g'} + \sum_{g'\neq g}^G\Sigma_s^{g'\to g} \varphi^{g'} -\left( -\div D^g \grad  + \Sigma_r^g \right)\varphi^g\right)}
{\sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\varphi^g\right)}
\label{eq:rmb}
\ee
\be
\frac{\bar{\beta}}{\Lambda} = \sum_{i=1}^I\frac{\bar{\beta}_i}{\Lambda} = 
\sum_{i=1}^I\frac{1}{\keff}\frac{\sum_{g=1}^G(\phi^{*g}, \chi_{d,i}^g\beta_i\sum_{g'=1}^G\nu^{g'} \Sigma_f^{g' }\varphi^{g'})}
{\sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\varphi^g\right)}
\label{eq:b}
\ee
\be
\bar{\lambda}_i = \frac{\sum_{g=1}^G(\phi^{*g},\chi_{d,i}^g\lambda_i C_i)}{\sum_{g=1}^G(\phi^{*g},\chi_{d,i}^gC_i)}
\label{eq:l}
\ee 
\end{subequations}
The following inner product definition has been used: $\left(\phi^{*g},f\right):=\int_D\phi^{*g}(\vec{r})f(\vec{r})d^3r$.

@@

Where $n$ is the time step, $k$ is the iteration number, and $\epsilon_{\varphi}$ is the numerical criteria provided by a user. Through testing, iterations do not always converge; in this circumstance, the maximum number of iterations is reached.

@@

Where $K$ is the IQS uniqueness expression:
\be 
K_n^{(k+1)} = \sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\varphi_n^{g,(k+1)}\right)
\label{eq:K}
\ee

@@

By far, the most common nonlinear system iteration method in MOOSE is Preconditioned Jacobian-Free Newton-Krylov with Generalized Minimal RESidual method (GMRES) as the linear system solver.  This section only describes the methods' application to IQS; a very detailed description of PJFNK is transcribed by Knoll in \cite{PJFNK_Knoll}. Essentially, the IQS system of equations can be described by \eqt{eq:IQSsys}.  Where $A$ is a matrix operator, $\varphi$ is the solution vector, and $F$ is the forcing vector.
\be
A(\varphi(p))\varphi=F(\varphi,p,t)
\label{eq:IQSsys}
\ee
Applying the residual based Newton-method yields:
\be
J\delta\varphi = -R(\varphi,p)
\ee 
Where $J$ is the Jacobian matrix defined as $J_{ij}=\partial R_i/\partial\varphi_j$, $\delta\varphi$ is the error in the iterating solution, and $R$ is the residual vector defined as $A\varphi-F$ (most methods simply define the residual instead of evaluating this expression).  For IQS, a residual evaluation entails a PRKE parameter evaluation and the PRKE evaluation over the entire macro-step.  Applying a preconditioner $P$ yields:
\be 
(JP^{-1})(P\delta\varphi)=-R(\varphi,p)
\ee
This resulting system can be split into two systems, represented by \eqts{eq:PJFNK_1}{eq:PJFNK_2}.
\be 
(JP^{-1})w = -R(\varphi,p)
\label{eq:PJFNK_1}
\ee
\be 
\delta\varphi = P^{-1}w
\label{eq:PJFNK_2}
\ee
Solving \eqt{eq:PJFNK_1} requires the evaluation of Jacobian operator, which is done each GMRES iteration in two steps:
\begin{enumerate}
\item Approximately solve the preconditioner system for $y$: $Py=w$
\item Perform matrix-free Jacobian operation: $Jy\approx[R(\varphi+\epsilon y,p')-R(\varphi,p)]/\epsilon$
\end{enumerate}
This Jacobian operation is a finite difference approach.  $\epsilon$ is a perturbation scalar and $p'$ is the amplitude computed with PRKE parameters calculated from the perturbed shape.  For IQS, the PRKE and its parameters must be evaluated for the perturbed and original system, although the unperturbed residual is stored from the beginning of the PJFNK iteration.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Predictor-Corrector version of IQS (IQS P-C)}
\label{sect:pc}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Temperature evaluation (\eqt{eq:temp}) is quite similar to the evaluation of the delayed neutron precursors from \sct{sect:dnp}. A typical implicit solver would simply use the interpolated flux at end of the temperature time step for the right hand side of the equation, a theta-method discretization.  However, IQS has much more information about the profile of the flux along the time step because of the micro-step amplitude evaluation.  Therefore, it is possible to solve for temperature using a semi-analytical approach, shown by \eqt{eq:temp_an}.
\be
T^{n+1} = T^n + \frac{\kappa_f}{\rho c_p} \sum_{g=1}^G\left(a_2(\Sigma_f^g\varphi^g)^{n+1} + a_1(\Sigma_f^g\varphi^g)^n\right)
\label{eq:temp_an}
\ee
Where $n$ corresponds to the beginning of the temperature step.  $a_1$ and $a_2$ are integration coefficients defined by \eqt{eq:a1} and \eqt{eq:a2}.  Any interpolation of the amplitude along the micro steps is possible for the integration, this application uses piece-wise linear.
\be
a_1 = \int_{t_n}^{t_{n+1}}\left(\frac{t_{n+1}-t'}{\Delta t}\right)p(t')dt'
\label{eq:a1}
\ee
\be
a_2 = \int_{t_n}^{t_{n+1}}\left(\frac{t'-t_n}{\Delta t}\right)p(t')dt'
\label{eq:a2}
\ee

@@

\begin{figure}[htpb!]
\centering
\begin{tikzpicture}[scale=2]
%Shape
\draw[] (0,2) -- (6,2) ;
\foreach \x in  {0,6}
\draw[shift={(\x,2)},color=black] (0pt,6pt) -- (0pt,-6pt);
\draw[shift={(0,2)},color=black] (0pt,0pt) -- (0pt,-6pt) node[below] {$t_n$};
\draw[shift={(6,2)},color=black] (0pt,0pt) -- (0pt,-6pt) node[below] {$t_{n+1}$};
\node(shape) at (-.25,2) {$\varphi:$};

@@

\end{tikzpicture}
\caption{Time scales and process of IQS with temperature feedback}
\label{fig:time}
\end{figure}

@@

\begin{figure}[!htpb]
\centering
\begin{tikzpicture}[every node/.style = {font=\normalsize}]

@@

\end{tikzpicture}
\caption{Visualization of fixed-point iteration and temperature update process for IQS}
\label{fig:proc}
\end{figure}

@@

A vital part of the verification and validation for IQS is analyzing error convergence.  Since any time discretization scheme can be applied to the shape equation of IQS, it is important to investigate IQS's performance to a variety of these schemes.  There is lack of literature that applies IQS to schemes other than implicit Euler; higher order schemes are never rigorously tested.  This research intends to apply a variety of schemes, including implicit Euler, Crank-Nicholson, backward difference formula (BDF), and diagonally implicit Runge-Kutta (DIRK), to test stability and error convergence.  For brevity in the following sections, the shape-diffusion equation is represented by a general operator notation, described by \eqt{eq:shape_mat}.
\be 
IV\frac{\partial \varphi}{\partial t} = A\varphi+b
\label{eq:shape_mat}
\ee
Where $IV$ is the inverse velocity operator, $A$ contains all the operations on $\varphi$ from the left-hand-side of \eqt{eq:shape}, and $b$ is the source from the precursors.

@@

A fairly simple way to evaluate the shape equation is to employ the $\theta$-scheme ($0\le\theta\le1$, explicit when $\theta=0$, implicit when $\theta=1$, and Crank-Nicholson when $\theta=1/2$) \cite{Ferziger}.  Generally, if there is a function $u$ whose governing equation is $\frac{du}{dt}=f(u,t)$, then the $\theta$-discretization is:
\be
\frac{u^{n+1}-u^n}{\Delta t}=(1-\theta)f(u^n,t) + \theta f(u^{n+1},t) \,.
\ee
Where $n$ is the previous time step and $n+1$ is the time step being evaluated. Applying this to \eqt{eq:shape_mat} yields:
\be
\varphi_{n+1} = (IV-\Delta t A_{n+1})^{-1}\left[\Delta t(1-\theta)(A_n\varphi_n+b_n) + \Delta t\theta b_{n+1} + IV\varphi_n\right]
\ee

@@

Singly-Diagonally-Implicit Runge-Kutta Method (SDIRK) is a powerful discretization method that involves solving the linear system in stages to reach a high order solution \cite{SDIRK}.  Generally, the method can be depicted by using a system where $dy/dt=f(t,y)$ and \eqt{eq:sdirk}.
\be 
y_{n+1} = y_n + \Delta t\sum_{i=1}^s b_ik_i
\label{eq:sdirk}
\ee
Where,
\be 
k_i = f(t_n+c_i\Delta t, y_n+\Delta t \sum_{j=1}^{i} a_{ij}k_j)
\ee
The coefficients $a_{ij}$, $b_i$, and $c_i$ can be represented by a Butcher tableau:
\begin{center}
\begin{tabular}{c|cccc}
$c_1$ & $a_{11}$ & &  & \\
'$c_2$ & $a_{21}$ & $a_{22}$ & & \\
$\vdots$ & $\vdots$ & $\vdots$ & $\ddots$ & \\
$c_s$ & $a_{s1}$ & $a_{s2}$ & $\ldots$ & $a_{ss}$\\
\hline
&$b_1$ & $b_2$ & $\ldots$ & $b_s$
\end{tabular}
\end{center}
An example used extensively in this research is SDIRK33, which is a third-order method with three stages.
\begin{center}
\begin{tabular}{c|ccc}
$\lambda$ & $\lambda$ & & \\
$\frac{1}{2}(1+\lambda)$ & $\frac{1}{2}(1-\lambda)$ & $\lambda$ & \\
$1$ & $\frac{1}{4}(-1+16\lambda-6\lambda^2)$ & $\frac{1}{4}(5-20\lambda+6\lambda^2)$ & $\lambda$\\
\hline
 & $\frac{1}{4}(-1+16\lambda-6\lambda^2)$ & $\frac{1}{4}(5-20\lambda+6\lambda^2)$ & $\lambda$
\end{tabular}
\end{center}
Where $\lambda\approx0.4358665215$ satisfies $1-9\lambda+18\lambda^2-6\lambda^3=0$.

@@

The relative error is computed as follows:
\be
e_n = \frac{\norm{\sum_{g=1}^G\varphi^g_{\Delta t/2} - \sum_{g=1}^G\varphi^g_{\Delta t}}}{\text{max}\left(\norm{\sum_{g=1}^G\varphi^g_{\Delta t/2}},\norm{\sum_{g=1}^G\varphi^g_{\Delta t}}\right)}
\label{eq:edt2}
\ee
If the error is smaller than the user-specified tolerance, $e_{tol}$, the time step is accepted. In addition, a new time step size is estimated as follows:
\be
\Delta t_{new} = S \Delta t \left(\frac{e_{tol}}{e_n}\right)^{\frac{1}{1+q}}
\label{eq:dt2}
\ee
Where $q$ is the convergence order of the time integration scheme being used and $S\simeq 0.8$ is a safety factor. If the error is larger than the user-specified tolerance, the time step is rejected. A new time step size is estimated using  \eqt{eq:edt2} as well. This process can be visualized by \figs{fig:dt2_1}{fig:dt2_2}.  Where a step involves a full convergence of shape, amplitude, and any multiphysics on the respective time step.

@@

\begin{figure}[!htpb]
\centering
\begin{tikzpicture}[every node/.style = {font=\large},scale=0.75]

@@

\end{tikzpicture}
\caption{Visualization of step doubling process with coding logic}
\label{fig:dt2_2}
\end{figure}

@@

A common precursor evaluation technique is using the theta method, described in \sct{sect:theta}. Applying this to \eqt{eq:prec}:
\be
\frac{C^{n+1}_i-C^n_i}{\Delta t}=(1-\theta)\beta_i S_f^np^n-(1-\theta)\lambda_i C^n_i + \theta\beta_i S_f^{n+1}p^{n+1}-\theta\lambda_i C^{n+1}_i
\ee
Where $S_f$ is the fission source equivalent for shape ($S_f^n=\sum_{g=1}^G(\nu\Sigma_f^g\varphi^g)^n$). Rearranging to solve for the precursor at the end of the time step yields
\be
C^{n+1}_i = \frac{1-(1-\theta)\Delta t\lambda_i}{1+\theta\Delta t\lambda_i}C^n_i + \frac{(1-\theta)\Delta t \beta_i}{1+\theta\Delta t\lambda_i}S_f^n p^n +  \frac{\theta\Delta t \beta_i}{1+\theta\Delta t\lambda_i}S_f^{n+1} p^{n+1}
\label{eq:dnp_theta}
\ee
Reporting this value of $C^{n+1}_i$, one can solve for the shape $\varphi^{n+1,g}$ as a function of $\varphi^{n,g}$ and $C^n_i$ (and $p^n$, $p^{n+1}$, $dp/dt|_n$ and  $dp/dt|_{n+1}$). Once $\varphi^{n+1,g}$ has been determined, $C^{n+1}_i$ is updated.  Applying this technique to implicit discretization is done by changing the definition the fission source ($S_f^n=\sum_{g=1}^G(\nu\Sigma_f^g\phi^g)^n$) and eliminating all $p$ terms.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{PRKE coefficients}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\chapter{\uppercase{Improved Quasi-Static Method}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory}
\label{sect:theory}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

To derive the amplitude equation, the shape/precursors equations are weighted by a space-dependent function and integrated over the phase-space. The weight function is typically the adjoint flux $\phi^{*g}$ at the initial time, which can be proven to minimize truncation error \cite{duderstadt1976nuclear}. The final expressions are given below:
\begin{subequations}
\be
\frac{dp}{dt}=\left[\frac{\rho-\bar{\beta}}{\Lambda}\right]p+\sum_{i=1}^I\bar{\lambda}_i\xi_i
\label{eq:p}
\ee
\be
\frac{d\xi_i}{dt}=\frac{\bar{\beta}_i}{\Lambda}-\bar{\lambda}_i\xi_i \quad 1 \le i \le I 
\label{eq:c}
\ee
\end{subequations}
This equation is also known as the point reactor kinetics equation (PRKE); where the reactivity, effective delayed-neutron fraction, and delayed-neutron precursor decay constant are defined as follows:
\begin{subequations}
\be
\frac{\rho-\bar{\beta}}{\Lambda} = 
\frac{ \sum_{g=1}^G\left(\phi^{*g},\frac{\chi_p^g}{\keff}\sum_{g'=1}^G \nu^{g'}_p \Sigma_f^{g'}\varphi^{g'} + \sum_{g'\neq g}^G\Sigma_s^{g'\to g} \varphi^{g'} +\left( \div D^g \grad  - \Sigma_r^g \right)\varphi^g\right)}
{\sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\varphi^g\right)}
\label{eq:rmb}
\ee
\be
\frac{\bar{\beta}}{\Lambda} = \sum_{i=1}^I\frac{\bar{\beta}_i}{\Lambda} = 
\sum_{i=1}^I\frac{1}{\keff}\frac{\sum_{g=1}^G(\phi^{*g}, \chi_{d,i}^g\beta_i\sum_{g'=1}^G\nu^{g'} \Sigma_f^{g' }\varphi^{g'})}
{\sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\varphi^g\right)}
\label{eq:b}
\ee
\be
\bar{\lambda}_i = \frac{\sum_{g=1}^G(\phi^{*g},\chi_{d,i}^g\lambda_i C_i)}{\sum_{g=1}^G(\phi^{*g},\chi_{d,i}^gC_i)}
\label{eq:l}
\ee 
\end{subequations}
The following inner product definition has been used: $\left(\phi^{*g},f\right):=\int_D\phi^{*g}(\vec{r})f(\vec{r})d^3r$.

@@

where $n$ is the time step, $k$ is the iteration number, and $\epsilon_{\varphi}$ is the numerical criteria provided by a user. Through testing, iterations do not always converge; in this circumstance, the maximum number of iterations is reached.

@@

where $K$ is the IQS uniqueness expression:
\be 
K_n^{(k+1)} = \sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\varphi_n^{g,(k+1)}\right)
\label{eq:K}
\ee

@@

By far, the most common nonlinear system iteration method in MOOSE is Preconditioned Jacobian-Free Newton-Krylov with Generalized Minimal RESidual method (GMRES) as the linear system solver.  This section only describes the methods' application to IQS; a very detailed description of PJFNK is transcribed by Knoll in \cite{PJFNK_Knoll}. Essentially, the IQS system of equations can be described by \eqt{eq:IQSsys}.  Where $A$ is a matrix operator, $\varphi$ is the solution vector, and $F$ is the forcing vector.
\be
A(\varphi(p))\varphi=F(\varphi,p,t)
\label{eq:IQSsys}
\ee
Applying the residual based Newton-method yields:
\be
J\delta\varphi = -R(\varphi,p)
\ee 
where $J$ is the Jacobian matrix defined as $J_{ij}=\partial R_i/\partial\varphi_j$, $\delta\varphi$ is the error in the iterating solution, and $R$ is the residual vector defined as $A\varphi-F$ (most methods simply define the residual instead of evaluating this expression).  For IQS, a residual evaluation entails a PRKE parameter evaluation and the PRKE evaluation over the entire macro-step.  Applying a preconditioner $P$ yields:
\be 
(JP^{-1})(P\delta\varphi)=-R(\varphi,p)
\ee
This resulting system can be split into two systems, represented by \eqts{eq:PJFNK_1}{eq:PJFNK_2}.
\be 
(JP^{-1})w = -R(\varphi,p)
\label{eq:PJFNK_1}
\ee
\be 
\delta\varphi = P^{-1}w
\label{eq:PJFNK_2}
\ee
Solving \eqt{eq:PJFNK_1} requires the evaluation of Jacobian operator, which is done each GMRES iteration in two steps:
\begin{enumerate}
\item Approximately solve the preconditioner system for $y$: $Py=w$
\item Perform matrix-free Jacobian operation: $Jy\approx[R(\varphi+\epsilon y,p')-R(\varphi,p)]/\epsilon$
\end{enumerate}
This Jacobian operation is a finite difference approach.  $\epsilon$ is a perturbation scalar and $p'$ is the amplitude computed with PRKE parameters calculated from the perturbed shape.  For IQS, the PRKE and its parameters must be evaluated for the perturbed and original system, although the unperturbed residual is stored from the beginning of the PJFNK iteration.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Predictor-Corrector IQS (IQS P-C)}
\label{sect:pc}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Temperature evaluation (\eqt{eq:temp}) is quite similar to the evaluation of the delayed neutron precursors from \sct{sect:dnp}. A typical implicit solver would simply use the interpolated flux at end of the temperature time step for the right hand side of the equation, a theta-method discretization.  However, IQS has much more information about the profile of the flux along the time step because of the micro-step amplitude evaluation.  Therefore, it is possible to solve for temperature using a semi-analytical approach, shown by \eqt{eq:temp_an}.
\be
T^{n+1} = T^n + \frac{\kappa_f}{\rho c_p} \sum_{g=1}^G\left(a_2(\Sigma_f^g\varphi^g)^{n+1} + a_1(\Sigma_f^g\varphi^g)^n\right)
\label{eq:temp_an}
\ee
where $n$ corresponds to the beginning of the temperature step.  $a_1$ and $a_2$ are integration coefficients defined by \eqt{eq:a1} and \eqt{eq:a2}.  Any interpolation of the amplitude along the micro steps is possible for the integration, this application uses piece-wise linear.
\be
a_1 = \int_{t_n}^{t_{n+1}}\left(\frac{t_{n+1}-t'}{\Delta t}\right)p(t')dt'
\label{eq:a1}
\ee
\be
a_2 = \int_{t_n}^{t_{n+1}}\left(\frac{t'-t_n}{\Delta t}\right)p(t')dt'
\label{eq:a2}
\ee

@@

\begin{figure}[htpb!]
\centering
\resizebox{\linewidth}{!}{
\begin{tikzpicture}[scale=2]
%Shape
\draw[] (0,2) -- (6,2) ;
\foreach \x in  {0,6}
\draw[shift={(\x,2)},color=black] (0pt,6pt) -- (0pt,-6pt);
\draw[shift={(0,2)},color=black] (0pt,0pt) -- (0pt,-6pt) node[below] {$t_n$};
\draw[shift={(6,2)},color=black] (0pt,0pt) -- (0pt,-6pt) node[below] {$t_{n+1}$};
\node(shape) at (-.25,2) {$\varphi:$};

@@

\end{tikzpicture}}
\caption{Time scales and process of IQS with temperature feedback}
\label{fig:time}
\end{figure}

@@

\begin{figure}[!htpb]
\centering
\resizebox{\linewidth}{!}{
\begin{tikzpicture}[every node/.style = {font=\normalsize}]

@@

\end{tikzpicture}}
\caption{Visualization of fixed-point iteration and temperature update process for IQS}
\label{fig:proc}
\end{figure}

@@

A vital part of the verification and validation for IQS is analyzing error convergence.  Since any time discretization scheme can be applied to the shape equation of IQS, it is important to investigate IQS's performance to a variety of these schemes.  There is lack of literature that applies IQS to schemes other than implicit Euler; higher order schemes are never rigorously tested.  This research intends to apply a variety of schemes, including implicit Euler, Crank-Nicholson, backward difference formula (BDF), and diagonally implicit Runge-Kutta (DIRK), to test stability and error convergence.  For brevity in the following sections, the shape-diffusion equation is represented by a general operator notation, described by \eqt{eq:shape_mat}.
\be 
IV\frac{\partial \varphi}{\partial t} = A\varphi+b
\label{eq:shape_mat}
\ee
where $IV$ is the inverse velocity operator, $A$ contains all the operations on $\varphi$ from the left-hand-side of \eqt{eq:shape}, and $b$ is the source from the precursors.

@@

A fairly simple way to evaluate the shape equation is to employ the $\theta$-scheme ($0\le\theta\le1$, explicit when $\theta=0$, implicit when $\theta=1$, and Crank-Nicholson when $\theta=1/2$) \cite{Ferziger}.  Generally, if there is a function $u$ whose governing equation is $\frac{du}{dt}=f(u,t)$, then the $\theta$-discretization is:
\be
\frac{u^{n+1}-u^n}{\Delta t}=(1-\theta)f(u^n,t) + \theta f(u^{n+1},t) \,.
\ee
where $n$ is the previous time step and $n+1$ is the time step being evaluated. Applying this to \eqt{eq:shape_mat} yields:
\be
\varphi_{n+1} = (IV-\Delta t A_{n+1})^{-1}\left[\Delta t(1-\theta)(A_n\varphi_n+b_n) + \Delta t\theta b_{n+1} + IV\varphi_n\right]
\ee

@@

Singly-Diagonally-Implicit Runge-Kutta Method (SDIRK) is a powerful discretization method that involves solving the linear system in stages to reach a high order solution \cite{SDIRK}.  Generally, the method can be depicted by using a system where $dy/dt=f(t,y)$ and \eqt{eq:sdirk}.
\be 
y_{n+1} = y_n + \Delta t\sum_{i=1}^s b_ik_i
\label{eq:sdirk}
\ee
where,
\be 
k_i = f(t_n+c_i\Delta t, y_n+\Delta t \sum_{j=1}^{i} a_{ij}k_j)
\ee
The coefficients $a_{ij}$, $b_i$, and $c_i$ can be represented by a Butcher tableau:
\begin{center}
\begin{tabular}{c|cccc}
$c_1$ & $a_{11}$ & &  & \\
'$c_2$ & $a_{21}$ & $a_{22}$ & & \\
$\vdots$ & $\vdots$ & $\vdots$ & $\ddots$ & \\
$c_s$ & $a_{s1}$ & $a_{s2}$ & $\ldots$ & $a_{ss}$\\
\hline
&$b_1$ & $b_2$ & $\ldots$ & $b_s$
\end{tabular}
\end{center}
An example used extensively in this research is SDIRK33, which is a third-order method with three stages.
\begin{center}
\begin{tabular}{c|ccc}
$\lambda$ & $\lambda$ & & \\
$\frac{1}{2}(1+\lambda)$ & $\frac{1}{2}(1-\lambda)$ & $\lambda$ & \\
$1$ & $\frac{1}{4}(-1+16\lambda-6\lambda^2)$ & $\frac{1}{4}(5-20\lambda+6\lambda^2)$ & $\lambda$\\
\hline
 & $\frac{1}{4}(-1+16\lambda-6\lambda^2)$ & $\frac{1}{4}(5-20\lambda+6\lambda^2)$ & $\lambda$
\end{tabular}
\end{center}
where $\lambda\approx0.4358665215$ satisfies $1-9\lambda+18\lambda^2-6\lambda^3=0$.

@@

The relative error is computed as follows:
\be
e_n = \frac{\norm{\sum_{g=1}^G\varphi^g_{\Delta t/2} - \sum_{g=1}^G\varphi^g_{\Delta t}}}{\text{max}\left(\norm{\sum_{g=1}^G\varphi^g_{\Delta t/2}},\norm{\sum_{g=1}^G\varphi^g_{\Delta t}}\right)}
\label{eq:edt2}
\ee
If the error is smaller than the user-specified tolerance, $e_{tol}$, the time step is accepted. In addition, a new time step size is estimated as follows:
\be
\Delta t_{new} = S \Delta t \left(\frac{e_{tol}}{e_n}\right)^{\frac{1}{1+q}}
\label{eq:dt2}
\ee
where $q$ is the convergence order of the time integration scheme being used and $S\simeq 0.8$ is a safety factor. If the error is larger than the user-specified tolerance, the time step is rejected. A new time step size is estimated using  \eqt{eq:edt2} as well. This process can be visualized by \figs{fig:dt2_1}{fig:dt2_2}.  Where a step involves a full convergence of shape, amplitude, and any multiphysics on the respective time step.

@@

\begin{figure}[!htpb]
\centering
\resizebox{\linewidth}{!}{
\begin{tikzpicture}[every node/.style = {font=\large},scale=0.75]

@@

\end{tikzpicture}}
\caption{Visualization of step doubling process with coding logic}
\label{fig:dt2_2}
\end{figure}

@@

A common precursor evaluation technique is using the theta method, described in \sct{sect:theta}. Applying this to \eqt{eq:prec}:
\be
\frac{C^{n+1}_i-C^n_i}{\Delta t}=(1-\theta)\beta_i S_f^np^n-(1-\theta)\lambda_i C^n_i + \theta\beta_i S_f^{n+1}p^{n+1}-\theta\lambda_i C^{n+1}_i
\ee
where $S_f$ is the fission source equivalent for shape ($S_f^n=\sum_{g=1}^G(\nu\Sigma_f^g\varphi^g)^n$). Rearranging to solve for the precursor at the end of the time step yields
\be
C^{n+1}_i = \frac{1-(1-\theta)\Delta t\lambda_i}{1+\theta\Delta t\lambda_i}C^n_i + \frac{(1-\theta)\Delta t \beta_i}{1+\theta\Delta t\lambda_i}S_f^n p^n +  \frac{\theta\Delta t \beta_i}{1+\theta\Delta t\lambda_i}S_f^{n+1} p^{n+1}
\label{eq:dnp_theta}
\ee
Reporting this value of $C^{n+1}_i$, one can solve for the shape $\varphi^{n+1,g}$ as a function of $\varphi^{n,g}$ and $C^n_i$ (and $p^n$, $p^{n+1}$, $dp/dt|_n$ and  $dp/dt|_{n+1}$). Once $\varphi^{n+1,g}$ has been determined, $C^{n+1}_i$ is updated.  Applying this technique to implicit discretization is done by changing the definition the fission source ($S_f^n=\sum_{g=1}^G(\nu\Sigma_f^g\phi^g)^n$) and eliminating all $p$ terms.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{PRKE Coefficients}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
